---
title: "Getting started with SIR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with SIR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 5,
  fig.height = 4
)
```

```{r}
library(SIR)
```

## A small synthetic example

We generate a small (6 by 6) dose-response matrix on a transformed scale,
with a known synergistic bump in the middle of the grid. Rows index doses
of drug A and columns doses of drug B.

```{r}
set.seed(2026)
I <- 6L; J <- 6L

# Monotone-decreasing single-agent effects, centered.
u <- -cumsum(abs(rnorm(I, mean = 0.30, sd = 0.10))); u <- u - mean(u)
v <- -cumsum(abs(rnorm(J, mean = 0.30, sd = 0.10))); v <- v - mean(v)

# Additive null surface.
Z_add <- outer(u, v, "+")

# Negative bump at the centre of the grid -> synergy under "decreasing" data.
ii <- seq_len(I); jj <- seq_len(J)
bump <- outer(ii, jj, function(i, j) {
  exp(-((i - (I + 1) / 2)^2 + (j - (J + 1) / 2)^2) / (2 * 1.2^2))
})
Z <- Z_add - 0.30 * bump + 0.05 * matrix(rnorm(I * J), I, J)
```

## Running the SIR test

The convenient entry point is `sir_test()`. It fits both the
monotone-additive null surface and a fully monotone two-dimensional
isotonic surface, then computes a wild-bootstrap p-value against the
additive null. We use a small bootstrap (`B = 100`) here so the vignette
runs quickly; in practice values around 200 to 500 are typical.

```{r}
fit <- sir_test(
  Z,
  B = 100,
  direction = "decreasing",
  synergy_sign = "negative",
  stat = "S2",
  seed = 1
)

fit
```

A more detailed view is provided by `summary()`:

```{r}
summary(fit)
```

## Inspecting the interaction surface

The fitted interaction surface `delta = theta_iso - theta_add` is
returned in `fit$delta`. Negative values indicate synergy under
viability-style data:

```{r}
round(fit$delta, 3)
```

A heatmap of the interaction surface is the default plot:

```{r}
plot(fit)
```

## Comparing against classical baselines

SIR also exposes the four most common synergy baselines, useful for
sanity-checking and methods comparisons. They consume long-format
dose-response tables:

```{r}
df <- data.frame(
  doseA = rep(seq_len(I), times = J),
  doseB = rep(seq_len(J), each = I),
  response = pmax(0, pmin(1, plogis(as.numeric(Z)))),
  replicate = "1"
)

bliss <- bliss_score(df, response_mode = "viability")
hsa   <- hsa_score(df, response_mode = "viability")

dim(bliss$synergy)
dim(hsa$synergy)
```

For full details on each helper, see `?additive_ordered_fit`,
`?isotonic_2d_fit`, `?wild_bootstrap`, and `?interaction_summaries`.
