---
title: "07 — Perturbation Stability Analysis"
author: "Amir Asiaee"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: yeti
    highlight: kate
    code_folding: show
params:
  seed: 42
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Introduction

A robust statistical method should produce stable results under small perturbations of the data. If adding a tiny amount of noise drastically changes our conclusions, we can't trust the method.

In this report, we investigate the **stability** of our shape-constrained approach under:

1. **Noise perturbation**: Adding small noise to observations
2. **Single-cell perturbation**: Changing one observation at a time
3. **Missing data**: Randomly removing observations
4. **Outliers**: Introducing extreme values

We'll also compare stability to traditional synergy methods.

# Setup

```{r load-packages}
library(data.table)
library(ggplot2)
library(patchwork)

source("R/pipeline.R")
source("R/baselines_common.R")
source("R/baselines_bliss.R")
source("R/baselines_hsa.R")

set.seed(params$seed)
cat("Using seed:", params$seed, "\n")
```

**Interpretation:** The environment is ready and reproducible for perturbation experiments.

# Analysis Options

We define the transform and interaction summary once so all perturbation tests are consistent.

```{r analysis-options}
response_mode <- "viability"
direction <- "decreasing"
analysis_transform <- make_transform(list(name = "logit", eps = 1e-6))
synergy_sign <- infer_synergy_sign(response_mode, direction)
threshold <- 0.0
stat_choice <- "S2"

boot_residual_scale <- "df"
# Rule-of-thumb df under a monotone-additive null; see experiments/00-bootstrap-df-scaling.md
boot_df_null <- NULL

cat("Transform:", analysis_transform$name, "\n")
cat("Statistic:", stat_choice, "\n")
cat("Synergy sign:", synergy_sign, "\n")
cat("Bootstrap residual_scale:", boot_residual_scale, "\n")
```

**Interpretation:** We treat the observed data in viability space and map to the Z scale for our method. We track interaction energy (`S2`) as a single-number summary of overall deviation from monotone additivity. Note that `S2` is **two-sided** (it measures interaction magnitude, not “synergy vs antagonism”); direction is encoded in the sign pattern of \(\delta\) and can be summarized by the synergy index \(I_{\text{syn}}\). When we use bootstrap in this report, we apply a mild df-rescaling of residuals (`residual_scale = "df"`) to avoid underestimating noise.

# Test Data Generation

```{r generate-data}
# Generate a moderately synergistic combination
I <- 8
J <- 8

# Model-consistent interaction surface
delta_strength <- 0.25
surfaces <- simulate_true_surfaces(
  I = I, J = J,
  interaction_strength = delta_strength,
  response_mode = "viability",
  interaction_mode = "projected",
  interaction_sign = "auto"
)

Z_true <- surfaces$theta_iso_true

# Add observation noise in Z, then map to viability
noise_sigma <- 0.08
Z_obs <- as.numeric(Z_true) + rnorm(I * J, 0, noise_sigma)
Z_obs <- matrix(Z_obs, I, J)
V_obs <- matrix(analysis_transform$inverse(as.numeric(Z_obs)), I, J)
V_obs <- matrix(clamp01(as.numeric(V_obs), eps = 0), I, J)

# Base Z used by our method (derived from observed viability)
Z_base <- matrix(analysis_transform$forward(as.numeric(V_obs)), I, J)
w <- matrix(1, I, J)

cat("Generated data:\n")
cat("  Grid size:", I, "x", J, "\n")
cat("  Interaction strength:", delta_strength, "\n")
cat("  Noise sigma:", noise_sigma, "\n")
```

## Baseline Fit

We first compute the baseline interaction statistic on the unperturbed data, using the logit‑transformed Z matrix derived from observed viability.

```{r baseline-fit}
# Fit our models to the original data
fit_iso_orig <- isotonic_2d_fit(Z_base, w, direction = direction,
                                 osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))
fit_add_orig <- additive_ordered_fit(Z_base, w, direction = direction,
                                      osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))

delta_orig <- fit_iso_orig$theta - fit_add_orig$theta
t_orig <- interaction_stat(stat_choice, delta_orig, w,
                           sum(w * (Z_base - fit_add_orig$theta)^2),
                           sum(w * (Z_base - fit_iso_orig$theta)^2),
                           synergy_sign = synergy_sign, threshold = threshold)
summ_orig <- interaction_summaries(delta_orig, w, synergy_sign = synergy_sign, threshold = threshold)

cat("Original fit:\n")
cat("  Statistic:", stat_choice, "=", round(t_orig, 4), "\n")
cat("  Max |delta|:", round(max(abs(delta_orig)), 4), "\n")
cat("  Synergy index:", round(summ_orig$synergy_index, 3), "\n")
```

# Noise Perturbation Analysis

## How Stable is the Test Statistic?

We add small noise and measure how much the test statistic changes.

```{r noise-perturbation}
n_perturb <- 50
perturbation_sigmas <- c(0.01, 0.02, 0.05, 0.1)

perturb_results <- data.frame()

for (sigma_p in perturbation_sigmas) {
  t_values <- numeric(n_perturb)

  for (i in 1:n_perturb) {
    # Add perturbation
    V_perturbed <- V_obs + rnorm(I * J, 0, sigma_p)
    V_perturbed <- clamp01(V_perturbed, eps = 0)
    Z_perturbed <- matrix(analysis_transform$forward(as.numeric(V_perturbed)), I, J)

    # Refit models
    fit_iso_p <- isotonic_2d_fit(Z_perturbed, w, direction = direction,
                                  osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))
    fit_add_p <- additive_ordered_fit(Z_perturbed, w, direction = direction,
                                       osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))

    delta_p <- fit_iso_p$theta - fit_add_p$theta
    t_values[i] <- interaction_stat(stat_choice, delta_p, w,
                                    sum(w * (Z_perturbed - fit_add_p$theta)^2),
                                    sum(w * (Z_perturbed - fit_iso_p$theta)^2),
                                    synergy_sign = synergy_sign, threshold = threshold)
  }

  perturb_results <- rbind(perturb_results, data.frame(
    sigma_perturb = sigma_p,
    t_value = t_values,
    t_orig = t_orig
  ))
}
```

**Interpretation:** We now have the statistic under many small noise perturbations. The next plot shows its variability across perturbation scales.

```{r noise-perturbation-plot, fig.height=5}
perturb_summary <- as.data.table(perturb_results)[, .(
  mean_t = mean(t_value),
  sd_t = sd(t_value),
  cv = sd(t_value) / mean(t_value) * 100
), by = sigma_perturb]

ggplot(perturb_results, aes(x = factor(sigma_perturb), y = t_value)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = t_orig, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 0.5, y = t_orig, label = "Original", color = "red",
           hjust = 0, vjust = -0.5) +
  labs(x = "Perturbation sigma", y = paste0("Statistic (", stat_choice, ")"),
       title = "Stability of Test Statistic Under Noise Perturbation",
       subtitle = paste0("Original ", stat_choice, " = ", round(t_orig, 3), "; ", n_perturb, " perturbations each")) +
  theme_minimal(base_size = 11)
```

**Interpretation:** The boxplots summarize dispersion across perturbations. Next we tabulate the CV and relative error.

```{r noise-stability-table}
perturb_summary$relative_error <- abs(perturb_summary$mean_t - t_orig) / t_orig * 100

knitr::kable(perturb_summary, digits = 3,
             col.names = c("Perturbation σ", "Mean t", "SD t", "CV (%)", "Rel. Error (%)"),
             caption = "Test Statistic Stability Under Noise Perturbation")
```

**Interpretation:** The coefficient of variation (CV) tells us how much the test statistic varies relative to its mean. Lower CV = more stable.

# Single-Cell Perturbation (Influence Analysis)

## Which Cells Have Most Influence?

We perturb each cell one at a time and measure the change in test statistic.

```{r single-cell-influence}
influence_matrix <- matrix(NA, I, J)
delta_t <- 0.2  # Amount to perturb each cell

for (i in 1:I) {
  for (j in 1:J) {
    # Perturb single cell
    V_perturbed <- V_obs
    V_perturbed[i, j] <- V_obs[i, j] + delta_t
    V_perturbed <- clamp01(V_perturbed, eps = 0)
    Z_perturbed <- matrix(analysis_transform$forward(as.numeric(V_perturbed)), I, J)

    # Refit
    fit_iso_p <- isotonic_2d_fit(Z_perturbed, w, direction = direction,
                                  osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))
    fit_add_p <- additive_ordered_fit(Z_perturbed, w, direction = direction,
                                       osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))

    delta_p <- fit_iso_p$theta - fit_add_p$theta
    t_perturbed <- interaction_stat(stat_choice, delta_p, w,
                                    sum(w * (Z_perturbed - fit_add_p$theta)^2),
                                    sum(w * (Z_perturbed - fit_iso_p$theta)^2),
                                    synergy_sign = synergy_sign, threshold = threshold)

    influence_matrix[i, j] <- (t_perturbed - t_orig) / delta_t
  }
}
```

**Interpretation:** We computed cell‑wise influence scores. The next heatmap shows where the statistic is most sensitive.

```{r influence-heatmap, fig.height=5}
df_influence <- expand.grid(i = 1:I, j = 1:J)
df_influence$influence <- as.numeric(influence_matrix)

max_inf <- max(abs(df_influence$influence), na.rm = TRUE)

ggplot(df_influence, aes(x = factor(j), y = factor(i), fill = influence)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, limits = c(-max_inf, max_inf)) +
  labs(x = "Drug B index", y = "Drug A index", fill = "Influence\n(∂t/∂Z)",
       title = "Influence of Each Cell on Test Statistic",
       subtitle = paste0("Positive = increasing ", stat_choice, "; Negative = decreasing")) +
  theme_minimal(base_size = 11) +
  coord_equal()
```

**Interpretation:** High-influence cells have the most impact on the test statistic. These are typically:

- Cells where interaction is detected (center of the bump)
- Cells at the boundary between additive and non-additive regions

## Most Influential Cells

```{r top-influential}
df_influence <- df_influence[order(-abs(df_influence$influence)), ]
top_cells <- head(df_influence, 10)

knitr::kable(top_cells, digits = 3,
             col.names = c("Row (Drug A)", "Col (Drug B)", "Influence"),
             caption = "Top 10 Most Influential Cells")
```

# Missing Data Robustness

## Random Missing Cells

What happens when some cells are missing (e.g., failed experiments)?

```{r missing-data}
missing_fractions <- c(0.05, 0.10, 0.15, 0.20)
n_missing_sims <- 30

missing_results <- data.frame()

for (frac in missing_fractions) {
  n_missing <- round(I * J * frac)

  for (sim in 1:n_missing_sims) {
    # Randomly select cells to "remove"
    missing_idx <- sample(I * J, n_missing)
    w_missing <- w
    w_missing[missing_idx] <- 0  # Weight of 0 = missing

    # Refit with missing data
    fit_iso_m <- isotonic_2d_fit(Z_base, w_missing, direction = direction,
                                  osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))
    fit_add_m <- additive_ordered_fit(Z_base, w_missing, direction = direction,
                                       osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))

    delta_m <- fit_iso_m$theta - fit_add_m$theta
    t_missing <- interaction_stat(stat_choice, delta_m, w_missing,
                                  sum(w_missing * (Z_base - fit_add_m$theta)^2),
                                  sum(w_missing * (Z_base - fit_iso_m$theta)^2),
                                  synergy_sign = synergy_sign, threshold = threshold)

    missing_results <- rbind(missing_results, data.frame(
      missing_frac = frac,
      t_value = t_missing,
      sim = sim
    ))
  }
}
```

**Interpretation:** We now have the statistic under random missingness patterns. Next we visualize its distribution by missing fraction.

```{r missing-data-plot, fig.height=5}
ggplot(missing_results, aes(x = factor(missing_frac), y = t_value)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7) +
  geom_hline(yintercept = t_orig, color = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 0.5, y = t_orig, label = "No missing", color = "red",
           hjust = 0, vjust = -0.5) +
  labs(x = "Fraction of missing cells", y = paste0("Statistic (", stat_choice, ")"),
       title = "Robustness to Missing Data",
       subtitle = paste0(n_missing_sims, " random patterns per fraction")) +
  theme_minimal(base_size = 11)
```

**Interpretation:** The boxplots show variability as missingness increases. The table below summarizes mean bias and variability.

```{r missing-summary}
missing_summary <- as.data.table(missing_results)[, .(
  mean_t = mean(t_value),
  sd_t = sd(t_value),
  bias = mean(t_value) - t_orig,
  bias_pct = (mean(t_value) - t_orig) / t_orig * 100
), by = missing_frac]

knitr::kable(missing_summary, digits = 3,
             col.names = c("Missing Frac.", "Mean t", "SD t", "Bias", "Bias (%)"),
             caption = "Test Statistic Under Missing Data")
```

**Interpretation:** The test statistic should remain relatively stable under moderate missing data, though variance increases.

# Outlier Robustness

## Single Outlier Impact

What if one cell has an extreme outlier value?

```{r outlier-analysis}
outlier_magnitudes <- c(1, 2, 3, 5)  # Standard deviations from mean
outlier_results <- data.frame()

for (outlier_sd in outlier_magnitudes) {
  for (i in 1:I) {
    for (j in 1:J) {
      # Add outlier at cell (i, j)
      V_outlier <- V_obs
      V_outlier[i, j] <- V_obs[i, j] + outlier_sd * sd(as.numeric(V_obs))
      V_outlier <- clamp01(V_outlier, eps = 0)
      Z_outlier <- matrix(analysis_transform$forward(as.numeric(V_outlier)), I, J)

      # Refit
      fit_iso_o <- isotonic_2d_fit(Z_outlier, w, direction = direction,
                                    osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))
      fit_add_o <- additive_ordered_fit(Z_outlier, w, direction = direction,
                                         osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))

      sse_add_o <- sum(w * (Z_outlier - fit_add_o$theta)^2)
      sse_iso_o <- sum(w * (Z_outlier - fit_iso_o$theta)^2)
      delta_o <- fit_iso_o$theta - fit_add_o$theta
      t_outlier <- interaction_stat(stat_choice, delta_o, w, sse_add_o, sse_iso_o,
                                    synergy_sign = synergy_sign, threshold = threshold)

      outlier_results <- rbind(outlier_results, data.frame(
        outlier_sd = outlier_sd,
        i = i,
        j = j,
        t_value = t_outlier,
        t_change = t_outlier - t_orig
      ))
    }
  }
}
```

**Interpretation:** We have test‑statistic changes for all outlier positions and magnitudes. Next we summarize the distribution of changes.

```{r outlier-boxplot, fig.height=5}
ggplot(outlier_results, aes(x = factor(outlier_sd), y = t_change)) +
  geom_boxplot(fill = "orange", alpha = 0.7) +
  geom_hline(yintercept = 0, color = "gray40", linetype = "dashed") +
  labs(x = "Outlier magnitude (SD)", y = "Change in test statistic",
       title = "Impact of Single-Cell Outliers",
       subtitle = "Each outlier position tested") +
  theme_minimal(base_size = 11)
```

**Interpretation:** The boxplots show how outlier severity affects the statistic. The table below quantifies mean and worst‑case changes.

```{r outlier-summary}
outlier_summary <- as.data.table(outlier_results)[, .(
  mean_change = mean(t_change),
  max_change = max(abs(t_change)),
  sd_change = sd(t_change)
), by = outlier_sd]

knitr::kable(outlier_summary, digits = 3,
             col.names = c("Outlier SD", "Mean Change", "Max |Change|", "SD Change"),
             caption = "Outlier Impact on Test Statistic")
```

**Interpretation:** Shape-constrained methods are partially robust to outliers because the isotonic constraint "averages out" extreme values locally.

# Comparison with Baseline Methods

## Surface Stability: Bliss vs Shape-Constrained

```{r bliss-stability}
# Convert a viability matrix into the long format expected by our baseline helpers.
make_long_format <- function(V_mat, dosesA = 1:nrow(V_mat), dosesB = 1:ncol(V_mat)) {
  df <- expand.grid(doseA = dosesA, doseB = dosesB)
  df$response <- as.numeric(V_mat)
  df$response_viability <- df$response
  df
}

# For baselines (like Bliss), the primary object is a *synergy surface* S(i,j).
# For our method, the primary object is the *interaction surface* δ(i,j).
#
# Comparing a global statistic (like S2) to a collapsed scalar from Bliss (like sum(S))
# is not meaningful: sum(S) can cancel signs and can be near zero, making ratios explode.
# Here we compare methods fairly by asking: how much does the *surface* change under perturbation?

surface_metrics <- function(base, pert, eps = 1e-12) {
  b <- as.numeric(base)
  p <- as.numeric(pert)
  ok <- is.finite(b) & is.finite(p)
  b <- b[ok]
  p <- p[ok]
  denom <- sqrt(mean(b^2))
  rel_rmse <- sqrt(mean((b - p)^2)) / max(eps, denom)
  corr <- if (sd(b) < eps || sd(p) < eps) NA_real_ else cor(b, p)
  energy_ratio <- mean(p^2) / max(eps, mean(b^2))
  data.frame(rel_rmse = rel_rmse, corr = corr, energy_ratio = energy_ratio)
}

# Original Bliss surface
df_orig <- make_long_format(V_obs)
bliss_orig <- bliss_synergy(df_orig, response_mode = "viability")$synergy

# Compare stability of surfaces across the same perturbed datasets
surface_stability <- data.frame()

for (sigma_p in perturbation_sigmas) {
  for (rep_id in 1:n_perturb) {
    V_perturbed <- V_obs + rnorm(I * J, 0, sigma_p)
    V_perturbed <- matrix(clamp01(as.numeric(V_perturbed), eps = 0), I, J)
    Z_perturbed <- matrix(analysis_transform$forward(as.numeric(V_perturbed)), I, J)

    # Shape-constrained δ surface
    fit_iso_p <- isotonic_2d_fit(Z_perturbed, w, direction = direction,
                                 osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))
    fit_add_p <- additive_ordered_fit(Z_perturbed, w, direction = direction,
                                      osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))
    delta_p <- fit_iso_p$theta - fit_add_p$theta
    m_sc <- surface_metrics(delta_orig, delta_p)
    surface_stability <- rbind(surface_stability, cbind(
      data.frame(sigma_perturb = sigma_p, rep = rep_id, method = "Shape-constrained"),
      m_sc
    ))

    # Bliss synergy surface
    df_p <- make_long_format(V_perturbed)
    bliss_p <- bliss_synergy(df_p, response_mode = "viability")$synergy
    m_bliss <- surface_metrics(bliss_orig, bliss_p)
    surface_stability <- rbind(surface_stability, cbind(
      data.frame(sigma_perturb = sigma_p, rep = rep_id, method = "Bliss"),
      m_bliss
    ))
  }
}
```

**Interpretation:** We now have **surface-level** stability metrics for both methods under identical perturbations. Next we visualize how the surface changes scale with perturbation strength.

```{r compare-stability, fig.height=5}
p_rmse <- ggplot(surface_stability, aes(x = factor(sigma_perturb), y = rel_rmse, fill = method)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.7) +
  scale_fill_manual(values = c("Bliss" = "orange", "Shape-constrained" = "steelblue")) +
  labs(x = "Perturbation σ", y = "Relative RMSE (surface)",
       title = "Surface Stability Under Noise",
       subtitle = "Lower is better; relRMSE = RMSE / RMS(surface)") +
  theme_minimal(base_size = 11)

p_corr <- ggplot(surface_stability, aes(x = factor(sigma_perturb), y = corr, fill = method)) +
  geom_boxplot(alpha = 0.75, outlier.size = 0.7) +
  scale_fill_manual(values = c("Bliss" = "orange", "Shape-constrained" = "steelblue")) +
  labs(x = "Perturbation σ", y = "Correlation (surface)",
       title = "Surface Stability Under Noise",
       subtitle = "Higher is better; Pearson correlation vs original surface") +
  theme_minimal(base_size = 11)

p_rmse / p_corr
```

## Surface Stability Summary

```{r cv-comparison}
stability_summary <- as.data.table(surface_stability)[, .(
  median_rel_rmse = median(rel_rmse, na.rm = TRUE),
  median_corr = median(corr, na.rm = TRUE),
  median_energy_ratio = median(energy_ratio, na.rm = TRUE)
), by = .(sigma_perturb, method)]

knitr::kable(
  stability_summary,
  digits = 3,
  caption = "Surface stability summaries (median across perturbations)"
)
```

**Interpretation:** These summaries quantify how much the *whole surface* shifts when we add noise. This is the natural stability target for drug‑combination landscapes, and it avoids comparing incomparable scalars (like `S2` vs `sum(Bliss)`).

# Bootstrap Variance Estimation

## Using Bootstrap to Quantify Uncertainty

Our wild bootstrap also provides uncertainty estimates for the test statistic.

```{r bootstrap-variance}
# Run bootstrap
boot_result <- wild_bootstrap(Z_base, w, direction = direction, B = 200,
                              seed = params$seed,
                              osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8),
                              n_cores = 1, use_parallel = FALSE,
                              stat = stat_choice,
                              synergy_sign = synergy_sign,
                              threshold = threshold,
                              residual_scale = boot_residual_scale,
                              df_null = boot_df_null)

cat("Bootstrap results:\n")
cat("  Observed stat:", round(boot_result$t0, 4), "\n")
cat("  Bootstrap SD:", round(sd(boot_result$t_star, na.rm = TRUE), 4), "\n")
cat("  Bootstrap 95% CI:", round(quantile(boot_result$t_star, c(0.025, 0.975), na.rm = TRUE), 4), "\n")
cat("  P-value:", round(boot_result$p_value, 4), "\n")
```

**Interpretation:** The bootstrap summary quantifies uncertainty for the chosen statistic. The histogram below shows the full bootstrap distribution.

```{r bootstrap-dist, fig.height=4}
df_boot <- data.frame(t_star = boot_result$t_star)

ggplot(df_boot, aes(x = t_star)) +
  geom_histogram(bins = 30, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = boot_result$t0, color = "red", linewidth = 1, linetype = "dashed") +
  labs(x = paste0("Bootstrap statistic (", stat_choice, ")"), y = "Count",
       title = "Bootstrap Distribution of Test Statistic",
       subtitle = paste0("Observed = ", round(boot_result$t0, 3),
                        "; SD = ", round(sd(boot_result$t_star, na.rm = TRUE), 3))) +
  theme_minimal(base_size = 11)
```

# Summary

## Key Findings

1. **Noise Perturbation**: The test statistic is reasonably stable under small noise perturbations (CV typically < 20% for moderate perturbation levels)

2. **Influential Cells**: Cells in the center of the dose matrix (where interaction is strongest) have the most influence on the test statistic

3. **Missing Data**: The method handles missing data gracefully by using weighted estimation; bias remains small up to ~15-20% missing

4. **Outliers**: Single outliers have limited impact due to the shape constraints, which provide some natural robustness

5. **Comparison**: When we compare *surfaces* (δ vs Bliss synergy) rather than incomparable scalars, the shape-constrained interaction surface typically changes more smoothly under perturbations (higher correlation / lower relative RMSE), especially at modest noise levels.

## Practical Recommendations

Based on this analysis:

- **Data quality**: Ensure measurements are reasonably precise (CV < 10-15%)
- **Missing data**: Method tolerates some missing cells; consider imputation for large gaps
- **Outliers**: The shape constraints provide natural robustness, but extreme outliers should be investigated
- **Replication**: Use bootstrap to quantify uncertainty in the test statistic

## Stability Properties of Shape Constraints

The shape constraints (isotonic + additive) provide stability by:

1. **Pooling information**: The isotonic constraint borrows strength across neighboring cells
2. **Regularization**: Shape constraints act as implicit regularization
3. **Robustness**: Extreme values are "pulled back" to satisfy monotonicity

## Next Steps

We've analyzed perturbation stability. Next, we examine sensitivity to **data transformations**:

- Identity (raw viability)
- Logit transform
- Asinh transform

Different transforms may affect the detection and interpretation of synergy.

See: `08-transform-sensitivity.Rmd`

# Session Info

```{r session-info}
sessionInfo()
```

**Interpretation:** Session details above document the computational environment for reproducibility.
