---
title: "08 — Transform Sensitivity Analysis"
author: "Amir Asiaee"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: yeti
    highlight: kate
    code_folding: show
params:
  seed: 42
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Introduction

Drug combination data can be analyzed on different scales:

- **Identity**: Raw viability values (0 to 1)
- **Logit**: $\text{logit}(y) = \log(y / (1-y))$ — maps (0,1) to (-∞, +∞)
- **Asinh**: $\text{asinh}(y) = \log(y + \sqrt{y^2 + 1})$ — similar to log but handles zeros
- **Log**: $\log(y)$ — for strictly positive data

The choice of transform affects:

1. How "additivity" is defined
2. The distribution of residuals
3. The detection of synergy/antagonism

In this report, we explore how different transforms affect our shape-constrained analysis.

# Setup

```{r load-packages}
library(data.table)
library(ggplot2)
library(patchwork)

source("R/pipeline.R")
source("R/transforms.R")

set.seed(params$seed)
cat("Using seed:", params$seed, "\n")
```

**Interpretation:** The transform utilities are loaded and the random seed is fixed for reproducibility.

# Analysis Options

We define the interaction summary and synergy direction once so all transform comparisons are on the same footing.

```{r analysis-options}
response_mode <- "viability"
direction <- "decreasing"
synergy_sign <- infer_synergy_sign(response_mode, direction)
threshold <- 0.0
stat_choice <- "S2"

boot_residual_scale <- "df"
# Rule-of-thumb df under a monotone-additive null; see experiments/00-bootstrap-df-scaling.md
boot_df_null <- NULL

cat("Statistic:", stat_choice, "\n")
cat("Synergy sign:", synergy_sign, "\n")
cat("Bootstrap residual_scale:", boot_residual_scale, "\n")
```

**Interpretation:** We use interaction energy (`S2`) as the default global interaction summary and keep it fixed across transforms. Because transforms change the geometry/scale of \(Z\), the *numeric magnitude* of `S2` can shift; the bootstrap p-value is still meaningful *within each transform* because it uses that transform’s fitted null. Note that `S2` is **two-sided** (interaction magnitude); direction is carried by the sign of \(\delta\) and summarized by the synergy index \(I_{\text{syn}}\). For calibration, we use df-rescaled residuals (`residual_scale = "df"`). If you want a scale-normalized alternative for cross-transform comparability, you can switch `stat_choice` to `S2_norm`.

# Understanding Transforms

## Available Transforms

```{r transform-definitions}
# Create transform objects
transform_identity <- make_transform(list(name = "identity"))
transform_logit <- make_transform(list(name = "logit", eps = 1e-4))
transform_asinh <- make_transform(list(name = "asinh", asinh_scale = 1))
transform_log <- make_transform(list(name = "log", eps = 1e-6))

transforms <- list(
  identity = transform_identity,
  logit = transform_logit,
  asinh = transform_asinh,
  log = transform_log
)

cat("Available transforms:\n")
for (name in names(transforms)) {
  cat("  -", name, "\n")
}
```

## Visualizing Transform Functions

```{r transform-curves, fig.height=5}
y_seq <- seq(0.01, 0.99, length.out = 100)

transform_df <- data.frame(
  y = rep(y_seq, 4),
  z = c(
    transform_identity$forward(y_seq),
    transform_logit$forward(y_seq),
    transform_asinh$forward(y_seq),
    transform_log$forward(y_seq)
  ),
  transform = rep(c("Identity", "Logit", "Asinh", "Log"), each = length(y_seq))
)

ggplot(transform_df, aes(x = y, y = z, color = transform)) +
  geom_line(linewidth = 1) +
  labs(x = "Original value (y)", y = "Transformed value (z)",
       title = "Transform Functions",
       subtitle = "How each transform maps viability values") +
  theme_minimal(base_size = 11) +
  scale_color_brewer(palette = "Set1")
```

**Key differences:**

- **Identity**: Linear, preserves original scale
- **Logit**: Expands tails, symmetric around y=0.5
- **Asinh**: Similar to log but stable near zero
- **Log**: Expands low values, compresses high values

## Why Transforms Matter for Synergy

The null hypothesis "no interaction" means different things under different transforms:

- **Identity**: $\theta_{ij} = \alpha + u_i + v_j$ (additive in viability)
- **Logit**: Additive in log-odds → multiplicative in odds
- **Log**: Additive in log → multiplicative in viability (Bliss-like)

```{r transform-interpretation}
cat("Interpretation of additivity:\n\n")
cat("Identity: V_combo = V_A + V_B - 1 (roughly)\n")
cat("  - Effects add linearly\n\n")
cat("Logit: logit(V_combo) = logit(V_A) + logit(V_B) - const\n")
cat("  - Odds multiply\n\n")
cat("Log: log(V_combo) = log(V_A) + log(V_B) + const\n")
cat("  - Viabilities multiply (Bliss independence)\n")
```

# Generate Test Data

## Creating Combination Data

```{r generate-data}
I <- 8
J <- 8

boot_df_null <- I + J - 3

# Generate a model-consistent true surface in Z-space (logit by default)
surfaces <- simulate_true_surfaces(
  I = I, J = J,
  interaction_strength = 0.2,
  response_mode = "viability",
  interaction_mode = "projected",
  interaction_sign = "auto"
)

Z_true <- surfaces$theta_iso_true

# Add noise in Z-scale, then map to viability
noise_sigma <- 0.05
Z_obs <- as.numeric(Z_true) + rnorm(I * J, 0, noise_sigma)

trans_logit <- make_transform(list(name = "logit", eps = 1e-6))
V_obs <- matrix(trans_logit$inverse(Z_obs), I, J)
V_obs <- matrix(clamp01(as.numeric(V_obs), eps = 0.02), I, J)

cat("Generated data:\n")
cat("  Range:", round(range(V_obs), 3), "\n")
cat("  Mean:", round(mean(V_obs), 3), "\n")
```

## Visualize Original Data

```{r viz-original, fig.height=4}
df_orig <- expand.grid(i = 1:I, j = 1:J)
df_orig$V <- as.numeric(V_obs)

ggplot(df_orig, aes(x = factor(j), y = factor(i), fill = V)) +
  geom_tile() +
  scale_fill_gradient2(low = "darkred", mid = "white", high = "darkblue",
                       midpoint = 0.5, limits = c(0, 1)) +
  labs(x = "Drug B index", y = "Drug A index", fill = "Viability",
       title = "Observed Viability Matrix") +
  theme_minimal(base_size = 11) +
  coord_equal()
```

# Applying Different Transforms

## Transform and Analyze

```{r apply-transforms}
results <- list()
w <- matrix(1, I, J)

for (trans_name in c("identity", "logit", "asinh")) {
  trans <- transforms[[trans_name]]

  # Transform data
  Z_trans <- matrix(trans$forward(as.numeric(V_obs)), I, J)

  # Fit models
  fit_iso <- isotonic_2d_fit(Z_trans, w, direction = direction,
                              osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))
  fit_add <- additive_ordered_fit(Z_trans, w, direction = direction,
                                   osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))

  # Interaction surface in transformed space
  delta_trans <- fit_iso$theta - fit_add$theta
  stat_value <- interaction_stat(stat_choice, delta_trans, w,
                                 sum(w * (Z_trans - fit_add$theta)^2),
                                 sum(w * (Z_trans - fit_iso$theta)^2),
                                 synergy_sign = synergy_sign, threshold = threshold)
  summ_trans <- interaction_summaries(delta_trans, w, synergy_sign = synergy_sign, threshold = threshold)

  # Bootstrap for p-value
  boot_result <- wild_bootstrap(Z_trans, w, direction = direction, B = 100,
                                seed = params$seed,
                                osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8),
                                n_cores = 1, use_parallel = FALSE,
                                stat = stat_choice,
                                synergy_sign = synergy_sign,
                                threshold = threshold,
                                residual_scale = boot_residual_scale,
                                df_null = boot_df_null)

  results[[trans_name]] <- list(
    transform = trans_name,
    Z_trans = Z_trans,
    theta_iso = fit_iso$theta,
    theta_add = fit_add$theta,
    delta = delta_trans,
    synergy_index = summ_trans$synergy_index,
    stat_value = stat_value,
    p_value = boot_result$p_value,
    boot_result = boot_result
  )

  cat(trans_name,
      ": stat =", round(stat_value, 4),
      ", p =", round(boot_result$p_value, 4),
      ", synergy_index =", round(summ_trans$synergy_index, 3), "\n")
}
```

**Interpretation:** We now have transform‑specific fits, statistics, and p‑values. The next table summarizes these side by side.

## Compare Results

```{r compare-results}
summary_df <- data.frame(
  Transform = names(results),
  stat_value = sapply(results, function(r) r$stat_value),
  p_value = sapply(results, function(r) r$p_value),
  synergy_index = sapply(results, function(r) r$synergy_index),
  max_delta = sapply(results, function(r) max(abs(r$delta))),
  mean_delta = sapply(results, function(r) mean(r$delta))
)

knitr::kable(summary_df, digits = 4,
             caption = paste0("Comparison Across Transforms (stat = ", stat_choice, ")"))
```

# Visualizing Transform Effects

## Transformed Data

```{r transformed-data-viz, fig.height=4}
plots_trans <- list()

for (trans_name in names(results)) {
  df <- expand.grid(i = 1:I, j = 1:J)
  df$value <- as.numeric(results[[trans_name]]$Z_trans)

  plots_trans[[trans_name]] <- ggplot(df, aes(x = factor(j), y = factor(i), fill = value)) +
    geom_tile() +
    scale_fill_viridis_c() +
    labs(title = paste0(trans_name, " transformed"),
         x = "", y = "", fill = "z") +
    theme_minimal(base_size = 9) +
    coord_equal() +
    theme(legend.position = "bottom")
}

wrap_plots(plots_trans, ncol = 3)
```

## Interaction Surfaces

```{r interaction-surfaces, fig.height=4}
plots_delta <- list()

for (trans_name in names(results)) {
  df <- expand.grid(i = 1:I, j = 1:J)
  df$delta <- as.numeric(results[[trans_name]]$delta)
  max_d <- max(abs(df$delta))

  plots_delta[[trans_name]] <- ggplot(df, aes(x = factor(j), y = factor(i), fill = delta)) +
    geom_tile() +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                         midpoint = 0, limits = c(-max_d, max_d)) +
    labs(title = paste0(trans_name, ": δ = θ_iso - θ_add"),
         x = "", y = "", fill = "δ") +
    theme_minimal(base_size = 9) +
    coord_equal() +
    theme(legend.position = "bottom")
}

wrap_plots(plots_delta, ncol = 3)
```

**Observation:** The interaction pattern is detected across all transforms, but the magnitude and shape may differ slightly.

## Residual Distributions

```{r residual-distributions, fig.height=5}
residual_df <- data.frame()

for (trans_name in names(results)) {
  Z <- results[[trans_name]]$Z_trans
  theta_iso <- results[[trans_name]]$theta_iso
  residuals <- as.numeric(Z - theta_iso)

  residual_df <- rbind(residual_df, data.frame(
    transform = trans_name,
    residual = residuals
  ))
}

ggplot(residual_df, aes(x = residual)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7) +
  facet_wrap(~ transform, scales = "free", ncol = 3) +
  labs(x = "Residual (Z - θ_iso)", y = "Count",
       title = "Residual Distributions by Transform") +
  theme_minimal(base_size = 10)
```

**Interpretation:** The residual histograms summarize distributional shape. Next we check normality more directly with Q–Q plots.

```{r residual-qq, fig.height=5}
ggplot(residual_df, aes(sample = residual)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  facet_wrap(~ transform, scales = "free", ncol = 3) +
  labs(title = "Q-Q Plots: Are Residuals Normal?",
       subtitle = "Points should follow red line if normal") +
  theme_minimal(base_size = 10)
```

**Interpretation:** Different transforms may produce different residual distributions. The logit transform often produces more symmetric residuals for viability data bounded in (0,1).

# Sensitivity to Transform Choice

## How Much Do Results Differ?

```{r transform-sensitivity}
# Compare pairwise correlation of delta surfaces
delta_matrices <- lapply(results, function(r) as.numeric(r$delta))
delta_cor <- cor(do.call(cbind, delta_matrices))
colnames(delta_cor) <- rownames(delta_cor) <- names(results)

cat("Correlation of interaction surfaces across transforms:\n")
print(round(delta_cor, 3))
```

**Interpretation:** The correlation matrix summarizes how similar the detected interaction surfaces are across transforms. Next we compare the resulting p-values.

```{r pvalue-comparison, fig.height=4}
# Plot p-values
pval_df <- data.frame(
  transform = names(results),
  p_value = sapply(results, function(r) r$p_value)
)

ggplot(pval_df, aes(x = transform, y = p_value, fill = transform)) +
  geom_col(alpha = 0.7) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "red") +
  annotate("text", x = 0.5, y = 0.055, label = "α = 0.05", color = "red", hjust = 0) +
  labs(x = "Transform", y = "P-value",
       title = "P-values Across Different Transforms") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")
```

## When Do Transforms Disagree?

```{r disagreement-analysis}
# Generate multiple datasets and check agreement
n_datasets <- 20
agreement_results <- data.frame()

set.seed(params$seed)

for (d in 1:n_datasets) {
  # Generate model-consistent data in logit space
  delta_strength <- runif(1, 0, 0.3)
  surfaces_d <- simulate_true_surfaces(
    I = I, J = J,
    interaction_strength = delta_strength,
    response_mode = response_mode,
    interaction_mode = "projected",
    interaction_sign = "auto"
  )

  Z_true_d <- surfaces_d$theta_iso_true
  Z_obs_d <- as.numeric(Z_true_d) + rnorm(I * J, 0, noise_sigma)
  V_obs_d <- matrix(transforms$logit$inverse(Z_obs_d), I, J)
  V_obs_d <- matrix(clamp01(as.numeric(V_obs_d), eps = 0.02), I, J)

  pvals <- c()
  for (trans_name in c("identity", "logit", "asinh")) {
    trans <- transforms[[trans_name]]
    Z_d <- matrix(trans$forward(as.numeric(V_obs_d)), I, J)

    boot_d <- wild_bootstrap(Z_d, w, direction = direction, B = 50,
                             seed = params$seed + d,
                             osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8),
                             n_cores = 1, use_parallel = FALSE,
                             stat = stat_choice,
                             synergy_sign = synergy_sign,
                             threshold = threshold,
                             residual_scale = boot_residual_scale,
                             df_null = boot_df_null)
    pvals[trans_name] <- boot_d$p_value
  }

  # Check agreement at α = 0.05
  reject <- pvals < 0.05
  all_agree <- all(reject) | all(!reject)

  agreement_results <- rbind(agreement_results, data.frame(
    dataset = d,
    delta_strength = delta_strength,
    p_identity = pvals["identity"],
    p_logit = pvals["logit"],
    p_asinh = pvals["asinh"],
    all_agree = all_agree
  ))
}
```

**Interpretation:** We now have p-values across transforms for multiple datasets. Next we summarize overall agreement.

```{r agreement-summary}
cat("Transform Agreement Analysis:\n")
cat("  Number of datasets:", n_datasets, "\n")
cat("  All transforms agree:", sum(agreement_results$all_agree), "/", n_datasets,
    "(", round(mean(agreement_results$all_agree) * 100, 1), "%)\n")
```

**Interpretation:** The agreement rate quantifies how often transform choice changes the accept/reject decision. The scatter plot below shows the pattern across effect sizes.

```{r agreement-plot, fig.height=5}
agreement_long <- melt(
  as.data.table(agreement_results),
  id.vars = c("dataset", "delta_strength", "all_agree"),
  measure.vars = c("p_identity", "p_logit", "p_asinh"),
  variable.name = "transform",
  value.name = "p_value"
)
agreement_long$transform <- gsub("p_", "", agreement_long$transform)

ggplot(agreement_long, aes(x = delta_strength, y = p_value, color = transform)) +
  geom_point(alpha = 0.6) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "gray40") +
  geom_smooth(method = "loess", se = FALSE, linewidth = 0.8) +
  labs(x = "True interaction strength", y = "P-value",
       title = "P-values vs True Interaction Strength",
       subtitle = "Different transforms; dashed line = α = 0.05") +
  theme_minimal(base_size = 11) +
  scale_color_brewer(palette = "Set1")
```

# Transform Selection Guidelines

## When to Use Each Transform

```{r transform-guidelines}
guidelines_df <- data.frame(
  Transform = c("Identity", "Logit", "Asinh", "Log"),
  `Best For` = c(
    "Viability data in [0,1]",
    "Bounded data, symmetric null",
    "Data with values near 0",
    "Strictly positive ratios"
  ),
  Pros = c(
    "Simple, interpretable",
    "Handles boundaries, symmetric residuals",
    "Stable at zero, smooth",
    "Multiplicative null (Bliss-like)"
  ),
  Cons = c(
    "Boundary effects",
    "Sensitive near 0 and 1",
    "Less common in literature",
    "Can't handle zeros"
  )
)

knitr::kable(guidelines_df, caption = "Transform Selection Guidelines")
```

## Recommendation

Based on typical drug combination data:

1. **Default choice**: **Logit** for viability data in (0,1)
   - Handles bounded nature of data
   - Symmetric treatment of low and high viability
   - Well-established in biostatistics

2. **When data has zeros**: **Asinh** or **Identity**
   - Asinh is stable at zero
   - Identity avoids transformation issues

3. **For multiplicative models**: **Log** transform
   - Aligns with Bliss independence assumptions
   - But requires strictly positive data

# Impact on Conclusions

## Summary Comparison

```{r final-comparison}
cat("Summary of transform effects on our test data:\n\n")

for (trans_name in names(results)) {
  r <- results[[trans_name]]
  conclusion <- if (r$p_value < 0.05) "REJECT null (interaction detected)" else "FAIL TO REJECT null"
  cat(sprintf("%-10s: %s = %6.3f, p = %.3f → %s\n",
              trans_name, stat_choice, r$stat_value, r$p_value, conclusion))
}
```

## Key Findings

1. **General agreement**: For moderate effect sizes, transforms usually agree on whether synergy is detected

2. **Edge cases**: Disagreement is more likely when:
   - Effect size is near the detection threshold
   - Data has many values near boundaries (0 or 1)
   - Sample size is small

3. **Transform affects interpretation**: "Additivity" means different things under different transforms

4. **Practical recommendation**: Report results for at least 2 transforms as sensitivity analysis

# Summary

## What We Learned

1. **Different transforms define different null hypotheses**:
   - Identity: additive in viability
   - Logit: multiplicative in odds
   - Log: multiplicative in viability (Bliss)

2. **Our shape-constrained test works with any transform**: Just apply the transform before fitting

3. **Results are usually robust** across transforms for clear synergy/antagonism

4. **Borderline cases may differ**: Use sensitivity analysis when effect size is small

## Recommendations

- **Report multiple transforms** for transparency
- **Default to logit** for standard viability data
- **Check residual normality** to guide transform choice
- **Document your choice** in publications

## Next Steps

We've analyzed synthetic data. Now let's apply our methods to **real public drug combination datasets**.

See: `09-public-data.Rmd`

# Session Info

```{r session-info}
sessionInfo()
```

**Interpretation:** Session details above document the computational environment for reproducibility.
