---
title: "10 — Summary: Shape-Constrained Drug Synergy Analysis"
author: "Amir Asiaee"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: yeti
    highlight: kate
    code_folding: show
params:
  seed: 42
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6
)

knitr::opts_knit$set(root.dir = normalizePath(".."))
```

# Overview

This document summarizes the key findings and methods from our exploration of **shape-constrained drug synergy analysis** across the previous nine reports.

## Report Series Recap

| Report | Topic | Key Contribution |
|--------|-------|------------------|
| 01 | Data Generation | Simulation framework, noise models, transforms |
| 02 | Isotonic 2D Fitting | Shape-constrained regression via QP |
| 03 | Additive Decomposition | Null hypothesis model $\theta = \alpha + u_i + v_j$ |
| 04 | Interaction Test | Wild bootstrap for p-value computation |
| 05 | Calibration & Power | Type I error control and power analysis |
| 06 | Baselines | Comparison with Bliss, HSA, Loewe, ZIP |
| 07 | Perturbation Stability | Robustness to noise, missing data, outliers |
| 08 | Transform Sensitivity | Effects of identity, logit, asinh transforms |
| 09 | Public Data | Application workflow for real datasets |

# Setup

```{r load-packages}
library(data.table)
library(ggplot2)
library(patchwork)

source("R/pipeline.R")

set.seed(params$seed)
```

**Interpretation:** Core functions are loaded and the seed is set for reproducibility of the summary examples.

# Analysis Options

We set the interaction summary and direction used in the examples below so readers can reproduce the exact statistic.

```{r analysis-options}
response_mode <- "viability"
direction <- "decreasing"
synergy_sign <- infer_synergy_sign(response_mode, direction)
threshold <- 0.0
stat_choice <- "S2"

boot_residual_scale <- "df"

cat("Statistic:", stat_choice, "\n")
cat("Synergy sign:", synergy_sign, "\n")
cat("Bootstrap residual_scale:", boot_residual_scale, "\n")
```

**Interpretation:** The summary examples below use `stat_choice = "S2"` (interaction energy) to quantify interaction globally. Like an \(F\)-statistic, `S2` is **two-sided**: it detects “interaction exists” but does not encode whether the interaction is synergy-like or antagonism-like. Direction comes from the sign pattern of the interaction surface \(\delta\); globally we summarize direction using **directional energies** \(S^2_-\) vs \(S^2_+\) (and a signed synergy index \(I_{\text{syn}}\)). For bootstrap calibration we use a mild df-based residual rescaling (`residual_scale = "df"`), which prevents the null bootstrap from understating the noise level.

# The Method in a Nutshell

## Core Idea

We frame drug synergy detection as a **hypothesis test**:

- **$H_0$ (Null)**: The combination effect is **monotone-additive**
  $$\theta_{ij} = \alpha + u_i + v_j$$
  where $u_1 \geq u_2 \geq \cdots \geq u_I$ and $v_1 \geq v_2 \geq \cdots \geq v_J$

- **$H_1$ (Alternative)**: The combination effect is **monotone** (but not necessarily additive)
  $$\theta_{ij} \text{ is monotone decreasing in both } i \text{ and } j$$

We define interaction in the **same geometry we test**:
\[
 \delta = \theta^{\text{iso}} - \mathrm{Proj}_{\mathcal{A}}(\theta^{\text{iso}}),
\]
where \(\mathrm{Proj}_{\mathcal{A}}\) is the monotone‑additive projection of a monotone surface. This guarantees the total surface is monotone and the interaction is the non‑additive component.

## Interaction Summaries (choose one for testing)

We can test multiple **global summaries** of the interaction surface \(\delta\). Common choices include:

- \(t_{\text{int}} = \text{SSE}_{\text{add}} - \text{SSE}_{\text{iso}}\) (fit improvement) and its studentized version \(t_{\text{int,norm}} = t_{\text{int}} / \text{SSE}_{\text{add}}\)
- \(S^2 = \sum w_{ij}\,\delta_{ij}^2\) (overall interaction energy) and its studentized version \(S^2_{\text{norm}} = S^2 / \text{SSE}_{\text{add}}\)
- \(S^2_- = \sum w_{ij}\max(-\delta_{ij}, 0)^2\) and \(S^2_+ = \sum w_{ij}\max(\delta_{ij}, 0)^2\) (directional energies)
- \(I_{\text{syn}} = (S^2_{\text{syn}}-S^2_{\text{ant}})/(S^2_{\text{syn}}+S^2_{\text{ant}})\in[-1,1]\) (signed direction index; \(+1\)=mostly synergy, \(-1\)=mostly antagonism)
- \(\max(-\delta)\) or \(\text{area}\{\delta < -c\}\) (local “any‑synergy” summaries)

In practice, statistics based directly on \(\delta\) (e.g., \(S^2\) and the directional energies \(S^2_-, S^2_+\)) are often more aligned with the interaction‑surface interpretation than SSE gaps, which can be noise‑dominated in strongly monotone settings. Studentized versions (e.g., \(S^2_{\text{norm}}\)) improve finite-sample calibration by removing scale sensitivity to the noise level.

## P-value via Wild Bootstrap

1. Fit additive model → get residuals $r_{ij}$
2. Generate bootstrap data: $Z_{ij}^* = \hat{\theta}_{ij}^{\text{add}} + \xi_{ij} \cdot r_{ij}$ where $\xi_{ij} \sim \text{Rademacher}$
3. Compute the chosen statistic \(T^*\) for each bootstrap sample
4. P-value = proportion of \(T^* \geq T^{\text{obs}}\)

# Visual Summary of the Method

```{r visual-summary, fig.height=8}
# Generate example data (model-consistent alternative)
I <- 8
J <- 8

set.seed(params$seed)
surfaces <- simulate_true_surfaces(
  I = I, J = J,
  interaction_strength = 0.25,
  response_mode = response_mode,
  interaction_mode = "projected",
  interaction_sign = "auto"
)

Z_true <- surfaces$theta_iso_true
Z_obs <- matrix(as.numeric(Z_true) + rnorm(I * J, 0, 0.08), I, J)
w <- matrix(1, I, J)

# Fit models
fit_iso <- isotonic_2d_fit(Z_obs, w, direction = direction,
                            osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))
fit_add <- additive_ordered_fit(Z_obs, w, direction = direction,
                                 osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8))

# Create visualization
plot_matrix <- function(mat, title, scale_type = "viridis") {
  df <- expand.grid(i = 1:nrow(mat), j = 1:ncol(mat))
  df$value <- as.numeric(mat)

  p <- ggplot(df, aes(x = factor(j), y = factor(i), fill = value)) +
    geom_tile() +
    labs(x = "", y = "", title = title, fill = "") +
    theme_minimal(base_size = 9) +
    coord_equal() +
    theme(legend.position = "bottom", legend.key.width = unit(1, "cm"))

  if (scale_type == "diverging") {
    max_abs <- max(abs(df$value))
    p <- p + scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                                  midpoint = 0, limits = c(-max_abs, max_abs))
  } else {
    p <- p + scale_fill_viridis_c()
  }
  p
}

p1 <- plot_matrix(Z_obs, "1. Observed Data (Z_obs)")
p2 <- plot_matrix(fit_add$theta, "2. Additive Fit (θ_add)")
p3 <- plot_matrix(fit_iso$theta, "3. Isotonic Fit (θ_iso)")
p4 <- plot_matrix(fit_iso$theta - fit_add$theta, "4. Interaction (δ = θ_iso - θ_add)", "diverging")

(p1 + p2) / (p3 + p4) +
  plot_annotation(title = "Shape-Constrained Drug Synergy Analysis: Visual Summary")
```

**Panel interpretation:**

1. **Observed data**: Raw dose-response matrix with noise
2. **Additive fit**: Best monotone-additive approximation (null model)
3. **Isotonic fit**: Best monotone approximation (alternative model)
4. **Interaction surface**: Difference reveals synergy (blue) or antagonism (red)

# Key Properties

## 1. Well-Calibrated Type I Error

```{r calibration-demo, fig.height=4}
# Quick calibration check
n_null_sims <- 30
B <- 50

pvals_null <- numeric(n_null_sims)

for (s in 1:n_null_sims) {
  # Generate null data (purely additive after projection)
  surfaces_null <- simulate_true_surfaces(
    I = I, J = J,
    interaction_strength = 0,
    response_mode = response_mode,
    interaction_mode = "projected",
    interaction_sign = "auto"
  )

  Z_null <- matrix(as.numeric(surfaces_null$theta_iso_true) + rnorm(I * J, 0, 0.1), I, J)

  df_null <- nrow(Z_null) + ncol(Z_null) - 3
  boot_s <- wild_bootstrap(Z_null, w, direction = direction, B = B,
                           seed = params$seed + s,
                           osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8),
                           n_cores = 1, use_parallel = FALSE,
                           stat = stat_choice,
                           synergy_sign = synergy_sign,
                           threshold = threshold,
                           residual_scale = boot_residual_scale,
                           df_null = df_null)
  pvals_null[s] <- boot_s$p_value
}

# ECDF plot
df_cal <- data.frame(p = sort(pvals_null), ecdf = (1:n_null_sims)/n_null_sims)

ggplot(df_cal, aes(x = p, y = ecdf)) +
  geom_abline(slope = 1, intercept = 0, color = "gray50", linetype = "dashed") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue") +
  labs(x = "P-value", y = "Empirical CDF",
       title = "Calibration Under the Null",
       subtitle = paste0(n_null_sims, " simulations; diagonal = perfect calibration")) +
  theme_minimal(base_size = 11) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1))
```

**Under the null hypothesis**, p-values should be approximately uniform (following the diagonal line). This indicates correct Type I error control.

## 2. Shape Constraints Are Natural

Drug combination responses are inherently monotone:

- **Higher dose → Equal or lower viability** (for cytotoxic drugs)
- This assumption is biologically justified
- The constraint provides regularization and robustness

## 3. Formal Hypothesis Testing

Unlike traditional synergy methods (Bliss, HSA, Loewe, ZIP) that only provide scores, our approach provides:

- Clear null and alternative hypotheses
- Formal test statistic
- Calibrated p-values
- Multiple testing correction support

# Comparison with Traditional Methods

```{r method-comparison}
comparison <- data.frame(
  Feature = c(
    "Null model",
    "Statistical test",
    "P-values",
    "Shape constraints",
    "Handles heteroskedasticity",
    "Computational cost"
  ),
  `Shape-Constrained` = c(
    "Monotone-additive",
    "Yes (bootstrap)",
    "Yes (calibrated)",
    "Yes (2D isotonic)",
    "Yes (weighted)",
    "Moderate-High"
  ),
  Bliss = c(
    "Multiplicative",
    "No*",
    "No*",
    "No",
    "No",
    "Low"
  ),
  HSA = c(
    "Max single-agent",
    "No*",
    "No*",
    "No",
    "No",
    "Low"
  ),
  Loewe = c(
    "Dose equivalence",
    "No*",
    "No*",
    "Partial",
    "No",
    "Low-Moderate"
  ),
  check.names = FALSE
)

knitr::kable(comparison, caption = "Method Comparison (*post-hoc tests sometimes added)")
```

# Practical Recommendations

## When to Use This Method

**Best suited for:**

- Rigorous statistical analysis of drug combinations
- Publication-quality results requiring p-values
- Studies where Type I error control is important
- Data with moderate-to-high quality (CV < 20%)

**Consider alternatives when:**

- Quick exploratory screening of many combinations
- Very noisy data where bootstrap is impractical
- Need for direct comparison with published Bliss/Loewe scores

## Recommended Workflow

```{r workflow}
cat("Recommended Analysis Workflow:\n\n")
cat("1. DATA PREPARATION\n")
cat("   - Quality check: remove failed wells, flag outliers\n")
cat("   - Normalize: relative to controls (DMSO/untreated)\n")
cat("   - Transform: apply logit for viability in (0,1)\n\n")

cat("2. MODEL FITTING\n")
cat("   - Fit isotonic model (shape-constrained)\n")
cat("   - Fit additive model (null hypothesis)\n")
cat("   - Compute chosen interaction statistic (e.g., S2 as default, S2_minus for synergy-direction energy, max_synergy)\n\n")

cat("3. STATISTICAL INFERENCE\n")
cat("   - Wild bootstrap (B = 500-1000 for publication)\n")
cat("   - Compute p-value with continuity correction\n")
cat("   - Multiple testing correction (FDR) if needed\n\n")

cat("4. INTERPRETATION\n")
cat("   - Examine interaction surface delta\n")
cat("   - Identify regions of synergy (delta < 0) vs antagonism (delta > 0)\n")
cat("   - Report effect sizes alongside p-values\n")
```

## Parameter Choices

```{r parameters}
params_df <- data.frame(
  Parameter = c("Transform", "Direction", "Statistic", "Bootstrap B", "OSQP tolerance", "Weight scheme"),
  Recommended = c("logit", "decreasing (viability)", "S2 (default)", "500-1000", "1e-8", "Uniform or 1/variance"),
  Notes = c(
    "Handles bounded (0,1) data well",
    "Higher dose = lower viability for cytotoxic",
    "Use df-rescaled wild bootstrap; S2_norm is optional for scale-normalization",
    "More = more precise p-values",
    "Tight tolerance for accurate solutions",
    "Uniform if no replicates; variance-weighted otherwise"
  )
)

knitr::kable(params_df, caption = "Recommended Parameter Settings")
```

# Limitations and Future Directions

## Current Limitations

1. **Computational cost**: Bootstrap is slower than simple synergy scores
2. **Grid structure**: Assumes regular dose-response matrix
3. **Monotonicity assumption**: May not hold for all drug types
4. **Sample size requirements**: Need reasonable grid (≥ 5×5)

## Future Directions

1. **Sparse/irregular grids**: Extend to non-rectangular designs
2. **Multi-drug combinations**: Beyond pairwise to 3+ drug combos
3. **Time-course data**: Incorporate temporal dynamics
4. **Bayesian extension**: Posterior distributions on interaction surface

# Reproducibility Checklist

For reproducible drug synergy analysis:

```{r checklist}
cat("REPRODUCIBILITY CHECKLIST\n")
cat("========================\n\n")
cat("[ ] Raw data archived with experiment metadata\n")
cat("[ ] Normalization procedure documented\n")
cat("[ ] Transform choice justified and recorded\n")
cat("[ ] Random seeds set for all stochastic steps\n")
cat("[ ] Bootstrap B and OSQP parameters recorded\n")
cat("[ ] Software versions captured (sessionInfo)\n")
cat("[ ] Code available in version control\n")
cat("[ ] Results validated on synthetic data first\n")
```

# Conclusions

## Main Takeaways

1. **Shape constraints are natural** for drug combination data and provide meaningful regularization

2. **Our approach provides formal hypothesis tests** with calibrated p-values, unlike traditional synergy scores

3. **Wild bootstrap handles heteroskedasticity** and non-standard null distributions

4. **The method is robust** to moderate noise, missing data, and outliers

5. **Transform choice matters** but results are usually consistent across transforms for clear effects

## Final Example

```{r final-example}
# Final demonstration
cat("=== Final Example Analysis ===\n\n")

# Use the data we generated earlier
delta_final <- fit_iso$theta - fit_add$theta
sse_add_final <- sum(w * (Z_obs - fit_add$theta)^2)
sse_iso_final <- sum(w * (Z_obs - fit_iso$theta)^2)
stat_final <- interaction_stat(stat_choice, delta_final, w, sse_add_final, sse_iso_final,
                               synergy_sign = synergy_sign, threshold = threshold)
summ_final <- interaction_summaries(delta_final, w, synergy_sign = synergy_sign, threshold = threshold)

boot_final <- wild_bootstrap(Z_obs, w, direction = direction, B = 100,
                             seed = params$seed,
                             osqp_pars = list(eps_abs = 1e-8, eps_rel = 1e-8),
                             n_cores = 1, use_parallel = FALSE,
                             stat = stat_choice,
                             synergy_sign = synergy_sign,
                             threshold = threshold,
                             residual_scale = boot_residual_scale,
                             df_null = nrow(Z_obs) + ncol(Z_obs) - 3)

cat("Test statistic (", stat_choice, "):", round(stat_final, 4), "\n")
cat("P-value:", round(boot_final$p_value, 4), "\n")
cat("Max interaction |delta|:", round(max(abs(fit_iso$theta - fit_add$theta)), 4), "\n")
cat("Synergy index:", round(summ_final$synergy_index, 3), "\n")

if (boot_final$p_value < 0.05) {
  cat("\nConclusion: REJECT null hypothesis\n")
  cat("Evidence of drug interaction detected!\n")
} else {
  cat("\nConclusion: FAIL TO REJECT null hypothesis\n")
  cat("No significant interaction detected.\n")
}
```

# Session Info

```{r session-info}
sessionInfo()
```

---

**Interpretation:** Session details above document the computational environment for reproducibility.

**Thank you for following this report series!**

For questions, issues, or contributions, please visit the project repository.
